Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2025-02-25 analysis/paper/paper.qmd
made some edits

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
deer_fami_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
deer_fami_NISP_per <- round(deer_fami_NISP/(total_NISP-unident), 3)*100
```

The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A). A similar trend also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B). Among the funal remains, deer family is the majority that accounts for about `r deer_fami_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts). Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer. Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temperoal sequences across 600 years. To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle. Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part throughout the cultural layers, followed by upper hindlimb and the less-meaty portions, such as lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The frequencies of axial elements, such as ribs, were generally low at Kiwulan, likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

@@

Surface modifications, including cut marks, burning damage, and fractures were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify). Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3. Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching. Some joints displyed chop marks along with cut marks (@fig-deer-cutmark), of which the clean, smooth cut surfaces and sharp marks suggest they were likly made by metal tools [@Greenfield1999]. @fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip. A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4. Not all types of joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult. Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation. In addition, cut marks on lower limbs can be associeted with skinning (@fig-deer-cutmark: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).

Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed on long bones across different layers (@fig-deer-fracture). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. This suggests more intensive marrow extraction in later periods. Chewing marks were also observed on some bones that show a series of shallow and smotth features (@fig-deer-fracture: e). In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-deer-fracture: c,d; @tbl-deer-modify).

@@

```{r}
#| label: fig-cutmark-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with cut marks, chop surfaces, and burning damage. Cut marks are indicated with red squares, and chop surfaces are indicated with black arrows. Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."

@@

```{r}
#| label: fig-fracture-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with fractures, natural modifications, and chewing marks. Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius with root etching, (d) femur shaft showing transverse fracture on the bottom, a percurssion scar in the middle with longitudinal split, and weathering, (e) scapula with chewing marks."

>>>>

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
deer_fami_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
deer_fami_NISP_per <- round(deer_fami_NISP/(total_NISP-unident), 3)*100
```

The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A). A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B). Among the faunal remains, the deer family is the majority accounting for about `r deer_fami_NISP_per`% of the bones identified to class level, followed by boar and cattle as the second and third taxon (@tbl-counts). Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer. Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temperoal sequences across 600 years. To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle. Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.

@@

Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases. Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges). The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot. The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991). The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences. A different pattern between layers is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter. In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.

@@

Surface modifications, including cut marks, burning damage, and fractures were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify). Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3. Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching. @fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip. A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4. Not all types of joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult. Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation. In addition, cut marks on lower limbs can be associeted with skinning (@fig-cutmark-img: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1). Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), along with narrow V-shaped cut marks observed on most bones, suggesting they were likely made with metal tools [@Greenfield1999].

Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4. Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img). These fractures indicate intentional marrow extraction at Kiwulan. CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5. Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023]. In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).

@@

```{r}
#| label: fig-cutmark-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with cut marks, chop surfaces, and burning damage. Cut marks are indicated by red squares, and chop surfaces are indicated by black arrows. Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."

@@

```{r}
#| label: fig-fracture-img
#| echo: false
#| fig-cap: "Examples of sika deer bones with fractures, natural modifications, and chewing marks. Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius showing tranverse fracture and root etching, (d) femur shaft showing longitudinal split, weathering, and a percurssion scar indicated by an arrow, (e) scapula with chewing marks."
